### 2015-01-14 engkaz-eamt-2015/engkaz-eamt-2015.tex
add art
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58326 72bbbca6-d526-0410-a7d9-f06f51895060

%
% File nodalida2015.tex
%
% Contact beata.megyesi@lingfil.uu.se
%
% Based on the instruction file for EACL 2014
% which in turn was based on the instruction files for previous 
% ACL and EACL conferences.

\documentclass[11pt]{article}
\usepackage[T2A,T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[russian,english]{babel}

\usepackage{times}
\usepackage{latexsym}
\usepackage{fixltx2e} %allows subscripts
%\usepackage{mathptmx}
%\usepackage{txfonts}
\usepackage{url}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

\usepackage{nodalida2015}

\usepackage{linguex}

\newcommand{\rus}[1]{\foreignlanguage{russian}{#1}}

\newcommand{\ft}[1]{\marginpar{\scriptsize F: #1}} % Fran's comments
\newcommand{\rr}[1]{\marginpar{\scriptsize R: #1}} % Rob's comments
\newcommand{\dm}[1]{\marginpar{\scriptsize D: #1}} % Detmar's comments
\newcommand{\lj}[1]{\marginpar{\scriptsize L: #1}} % Laura's comments

\title{Automatic word stress annotation of Russian running text (max 8 pages)}

\author{Author1 \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain.com} \\\And
  Author2 \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain.com} \\}

\date{2015}

\begin{document}
\maketitle
\begin{abstract}
  We evaluate the effectiveness of tools we developed for automatically annotating 
  word stress in Russian running text. Given an adequate lexicon with specified stress, the
  primary obstacle for correct stress placement is disambiguating homographic forms. Using a
  Constraint Grammar, we achieve a X\% improvement over the baseline undisambiguated performance.
\end{abstract}

%\section{Credits}
%
%The template had a credits section for the history of the template, but we don't need
%a Credits section, right?

@@

Russian word stress is variable. The inflecting word classes include complex patterns of
shifting stress, and a lexeme's stress pattern cannot be predicted from surface forms.
Although lexical stress and its attendant vowel reduction are a prominent feature of
spoken Russian, standard written Russian does not typically mark word stress.\footnote{
Texts intended for native speakers sometimes mark stress on words that cannot be 
disambiguated through context, although this is somewhat rare.} Without information about lexical 
stress position, 
correctly converting written Russian text to speech is impossible. This is true both for humans
(e.g. foreign language students) and computers (e.g. text-to-speech).

We identify three different types of word stress ambiguity. \emph{Intraparadigmatic}
stress ambiguity refers
to homographic wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

\ex. Intraparadigmatic homographs \label{ex:intrahom}
\a. \rus{т\'{е}ла} \emph{t\'{e}la} `body.\textsubscript{SG-GEN}' \label{ex:bodySGGEN}
\b. \rus{тел\'{а}} \emph{tel\'{a}} `body.\textsubscript{PL-NOM}' \label{ex:bodyPLNOM}

The remaining two types of stress ambiguity occur between lexemes. \emph{Morphosyntactically
congruent} stress ambiguity occurs between homographs that belong to separate lexemes, 
and whose morphosyntactic values are identical, as shown in Table~\ref{table:MScongruent}.

\begin{table}
\begin{center}
\begin{tabular}{c|ll}
\bf Number-Case & \bf `castle' & \bf `lock' \\
\hline
\texttt{SG-NOM} & \rus{з\'{a}мок} & \rus{зам\'{о}к} \\
\texttt{SG-GEN} & \rus{з\'{a}мка} & \rus{замк\'{а}} \\
\texttt{PL-NOM} & \rus{з\'{a}мки} & \rus{замк\'{и}} \\
\texttt{PL-GEN} & \rus{з\'{a}мков} & \rus{замк\'{о}в} \\
... & \hskip 1em ... & \hskip 1em ... \\
\end{tabular}
\end{center}
\caption{ \label{table:MScongruent} Morphosyntactically congruent homographs. }
\end{table}

\emph{Morphosyntactically incongruent} stress ambiguity occurs between homographs
that belong to separate lexemes, and whose morphosyntactic values are different, as shown in 
\ref{ex:MSincongruent}.

\ex. Morphosyntactically incongruent homographs \label{ex:MSincongruent}
\a. \rus{попад\'{а}ть} \emph{popadát'} `fall.\textsubscript{INF-IPFV}' \\
    \rus{поп\'{а}дать} \emph{popádat'} `fall.\textsubscript{INF-PFV}'
\b. \rus{узнаёт} \emph{uznaët} `finds out.\textsubscript{PRS-IPFV-3SG}' \\
    \rus{узн\'{а}ет} \emph{uznáet} `finds out.\textsubscript{FUT-PFV-3SG}'
\c. \rus{н\'{а}шей} \emph{nášej} `our.\textsubscript{F-SG-GEN/DAT/LOC...}'\\
    \rus{наш\'{е}й} \emph{našéj} `sew on.\textsubscript{IMP-2SG}'
\d. \rus{дор\'{о}га} \emph{doróga} `road.\textsubscript{N-F-SG-NOM}'\\
    \rus{дорог\'{а}} \emph{dorogá} `dear.\textsubscript{ADJ-F-SG-PRED}'

\rr{another relevant issue is whether the ambiguity can be resolved at all. Maybe this is 
something we should bring up in the discussion section. For example, when is stress marked in 
authentic by-native-for-native texts? A computer should not be expected to predict the stress on 
such words (although a computer could possibly identify words that should be marked for 
natives.)} It should be noted that not all morphosyntactic ambiguity leads to stress ambiguity. For example, all inflecting modifiers have the same form for F-SG-GEN, F-SG-LOC, F-SG-DAT and 
F-SG-INS. We refer to this as \emph{stress-irrelevant} ambiguity, since all readings have
the same stress placement. Stress-irrelevant ambiguity can be intraparadigmatic and/or interlexical.

Different types of stress ambiguity are overcome by different information. Detailed 
part-of-speech tagging can determine the stress of intraparadigmatic homographs and
morphosyntactically incongruent homographs, whereas word sense disambiguation can 
determine the stress position of morphosyntactically congruent homographs. In the case of
running text in Russian, most stress placement ambiguity is rooted in morphosyntactic
ambiguity, so in this study we focus on the role of part-of-speech tagging in improving
automatic stress placement, leaving word sense disambiguation to future work.

\subsection{Background and Task Definition}

Automatic stress placement in Russian is similar to Diacritic Restoration, a task
which has received increasing interest over the last 20 years. Generally speaking, diacritics
disambiguate otherwise homographic wordforms, so missing diacritics can
complicate many NLP tasks, such as text-to-speech. For example, speakers 
of Czech may type emails and other communications
without standard diacritics. In order to generate speech from these texts, they must first be 
normalized by restoring diacritics. A slightly different
situation arises with Arabic and Hebrew, since standard orthography lacks vowel diacritics
altogether. For such languages, the `restoration' of vowel diacritics results in less
ambiguity than in standard orthography. For languages with inherently ambiguous orthography,
it may be more precise to refer to this as `diacritic enhancement', since it produces
text that is less ambiguous than the standard language. In this sense, Russian orthography
is similar to Arabic and Hebrew, since its vowel qualities are underspecified in
standard orthography.

Many studies of Russian text-to-speech and automatic speech recognition make note of the 
difficulties caused by the shortcomings of their stress-marking resources (e.g. CITE Krivnova 
1998), but very few studies have targeted stress-marking itself, and they tend to
focus on placing stress on unknown words, with almost no attention to the problems that
arise with known words. For example, 
\cite{Xomicevic_2008} developed a set of heuristics for guessing stress placement on 
unknown words in Russian. More recently, CITE Hall \& Sproat (2013) trained a maximum entropy
model on a dictionary of Russian words, achieving 83.9\% accuracy for stress on unknown words 
and 98.7\% on known words. One major shortcoming of evaluating on wordlists is 
that they do not reflect wordform frequency.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms\rr{might it be worth computing the difference in stress
ambiguity between wordform lists and running text? i.e. what percentage of tokens have ambiguous
stress.}, so evaluations based on wordlists - as opposed to running
text - have inflated results. Furthermore, working with running text allows for the
possibility of disambiguating homographs based on syntactic context.

So far, the implicit target application the few studies related to automatic stress placement
in Russian has been text-to-speech and automatic speech recognition. However, the target 
application of our stress
annotator is a different domain: language learning. Since standard Russian does not contain 
stress-markings, learners are frequently unable to pronounce unknown words correctly without
referencing a dictionary or similar resources. In the context
of language learning, marking stress incorrectly is arguably worse than not marking it
at all. Because of this, we want our stress annotator to be able to abstain from marking
stress on words that it is unable to resolve with high confidence.

\begin{table*}[t]
  \centering
  \begin{tabular}{r|c|c}
    & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} \\
    & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} \\
    & \rus{костить}-V-IPFV-IMP\hskip 1em\rus{кост\'{и}} & \\
    \hline
    {\small {\tt bare}} & \rus{кости} & \rus{кости} \\
    {\small {\tt safe}} & \rus{кости} & \rus{к\'{о}сти} \\
    {\small {\tt guess}} & \rus{к\'{о}сти} or \rus{кост\'{и}} & \rus{к\'{о}сти}\\
    {\small {\tt guessFreq}} & \rus{к\'{о}сти} & \rus{к\'{о}сти}
  \end{tabular}
  \caption{Example output of each stress placement approach, given a particular set of readings}
  \label{tab:conditions}
\end{table*}

\subsection{Stress Corpus}

Russian texts with marked word stress are relatively rare, except in materials
for language learners, which are predominantly proprietary. Our corpus was collected from
free texts on Russian language-learning websites.\footnote{For a list of URL's, please 
contact the authors.} This small corpus (N=7689) is representative of texts 
that learners of Russian are likely to encounter in their studies. These texts 
include excerpts from well-known literary works, as well as dialogs and 
prose and individual sentences that were written for learners.

Unfortunately, the general practice for marking stress in Russian is to
\emph{not} mark stress on monosyllabic tokens, effectively assuming that
all monosyllabics are stressed. However, this approach 
is not well-motivated. Many words -- both monosyllabic and multisyllabic -- are 
unstressed, especially among prepositions, conjunctions, and particles.
\rr{multisyllabics that can be unstressed: \rus{через, между}} Furthermore, there are many
high-frequency monosyllabic homographs that can be either stressed or unstressed, 
depending on their part of speech, or particular collocations. Clearly, for
such words, one cannot assume that they are stressed on the basis of their syllable count.

Based on these considerations, we built our tools to mark stress without regard to
the number of syllables in a word. However, because our gold corpus texts do not
mark stress on monosyllabic words, we cannot evaluate our annotation of those words.

\section{Automatic Stress Placement}

State-of-the-art morphological analysis in Russian is based on finite-state
technology.\footnote{add footnote or citations for Dialog, mystem, etc.} To our knowledge,
no open-source resources are available for analyzing and generating stressed wordforms.
Therefore, we developed free and open-source finite-state tools capable of analyzing and
generating stressed wordforms. A finite-state transducer\footnote{Using two-level 
morphology, xfst/hfst, blah, blah, blah} (FST) generates all possible morphosyntactic readings
of each wordform, and a constraint grammar\footnote{vislcg3, blah blah} then removes
some extra readings based on syntactic context. The ultimate success of our stress placement
system depends on the performance of the Constraint Grammar. Ideally, the Constraint Grammar would
successfully remove all but the correct reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

The {\small {\tt bare}} approach is to not mark stress on words with more than one reading.
The {\small {\tt safe}} approach is to mark stress only on
tokens whose morphosyntactic ambiguity is stress-irrelevant. The {\small {\tt guess}}
approach is to randomly select one of the available readings, and abstain if no readings
are available. The {\small {\tt guessFreq}} approach is to select the reading that is most frequent.
\footnote{Frequency data taken from a separate hand-disambiguated corpus.} If none of the readings is found in the corpus, then it selects the reading
with the tag sequence (lemma removed) that is most frequent in our corpus. If the tag sequence is not found
in our corpus, then it backs off to the  {\small {\tt guess}} algorithm. Note that for tokens with
stress-irrelevant ambiguity, {\small {\tt guess}} and {\small {\tt guessFreq}} produce
the same result as the {\small {\tt safe}} method.

So far, the approaches discussed are
dependent on the availability of readings from the FST. Two final approaches address unknown tokens.
The {\small {\tt guessSyll}} and {\small {\tt guessFreqSyll}} approaches are identical to 
{\small {\tt guess}} and {\small {\tt guessFreq}}, respectively,
except that for unknown tokens, they places the stress near the end of the stem.\footnote{Determining 
which vowel is stem-final
is itself a difficult task. Our algorithm merely places the stress on the last vowel that is 
followed by a consonant.} \rr{Explain why stem-final is a good guess (and cite someone?)}
\rr{Discuss the success of the guessSyll approach in the discussion section.}

For our baseline, we take the output of our morphological analyzer (without the Constraint 
Grammar) in combination with the {\small {\tt bare}}, {\small {\tt safe}}, 
{\small {\tt guess}} and {\small {\tt guessFreq}} approaches.

We also compare our outcomes with the RussianGram\footnote{http://russiangram.com/} plugin 
for the Google Chrome web browser. RussianGram is not open-source, so we
can only guess what technologies support the service. In any case, it provides a
meaningful reference point for the success of each of the methods described above.

\section{Results}

\begin{table*}[t]
  \centering
  \begin{tabular}{r | r r r | r r }
    approach & correct\% & error\% & abstention\% & totTry\% & totFail\% \\
    \hline
    \hline
    {\small {\tt noCGbare}} & 30.43 & 0.17 & 69.39 & 30.61 & 69.57 \\
    {\small {\tt noCGsafe}} & 90.07 & 0.49 & 9.44 & 90.56 & 9.93 \\
    {\small {\tt noCGguess}} & 94.34 & 3.36 & 2.30 & 97.70 & 5.66 \\
    {\small {\tt noCGguessFreq}} & 95.53 & 2.59 & 1.88 & 98.12 & 4.47 \\
    {\small {\tt noCGguessSyll}} & 94.99 & 4.05 & 0.96 & 99.04 & 5.01 \\
    {\small {\tt noCGguessFreqSyll}} & 95.83 & 3.46 & 0.72 & 99.28 & 4.17 \\
    \hline
    {\small {\tt CGbare}} & 45.78 & 0.44 & 53.78 & 46.22 & 54.22 \\
    {\small {\tt CGsafe}} & 93.21 & 0.74 & 6.05 & 93.95 & 6.79 \\
    {\small {\tt CGguess}} & 95.50 & 2.59 & 1.90 & 98.10 & 4.50 \\
    {\small {\tt CGguessFreq}} & 95.73 & 2.40 & 1.88 & 98.12 & 4.27 \\
    {\small {\tt CGguessSyll}} & 95.92 & 3.33 & 0.74 & 99.26 & 4.08 \\
    {\small {\tt CGguessFreqSyll}} & 96.15 & 3.14 & 0.72 & 99.28 & 3.85 \\
    \hline
    {\small {\tt RussianGram}} & 90.09 & 0.79 & 9.12 & 90.88 & 9.91
  \end{tabular}
  \caption{Results of stress placement task evaluation. (N = 4048)}
  \label{tab:results}
\end{table*}

Single-syllable words and words that are unstressed in the gold standard were
not evaluated for reasons discussed above, leaving 4048 tokens for evaluation.
Recall was computed by testing whether stressed 
multisyllabic wordforms in the gold standard stress corpus were found in the 
finite-state transducer. Out of 4048 tokens, 3949 were found in the FST, which is equal to 97.55\%.
This number represents the ceiling for rule-based methods relying on the FST.
Higher scores are only achievable by expanding the FST or by using syllable-guessing algorithms.
After running the Constraint Grammar, 3937 tokens were found in the FST, which
indicates that the Constraint Grammar ruled out 0.20\% of the correct forms.
By inspecting these errors, we can improve the Constraint Grammar.

Results were compiled for each of the nine conditions discussed above: 
\emph{without} the Constraint Grammar ({\small {\tt noCG}}) X 4 approaches, 
\emph{with} the Constraint Grammar ({\small {\tt CG}}) X 4 approaches, 
and RussianGram ({\small {\tt RG}}). 
Failures to correctly place stress were split into two categories: errors and abstentions. 
If the stress tool outputs a stressed wordform, and it is incorrect, then it is
counted as an `error'. If the stress tool outputs an unstressed wordform, then
it is counted as an `abstention'. The sum of the errors and abstentions is referred
to as `totFail'. The sum of correct attempts and `errors' is referred to as `totTry'.
Results are given in Table~\ref{tab:results}.

WE NEED TO COMPUTE SIGNIFICANCE ON THESE RESULTS

\section{Discussion}

%\section*{Acknowledgments}
%
%Do not number the acknowledgment section. Do not include this section
%when submitting your paper for review.

\bibliographystyle{acl}
\bibliography{rusaccent}

>>>>

\documentclass[a4paper,11pt]{article}

\usepackage{url}

\begin{document}

@@
